### 2022-11-28 dissertation/chapters/07-methodology.tex
🚧 (final-review): Chapter 7 WIP

Each empirical tests starts with a question to be answered. A hypothesis is formulated for the outcome of the empirical test. The empirical tests are then designed around the implementation of the components under evaluation. Each empirical test defines the configuration of elements and generally include a set of parameters that are altered between experiments. Each experiment is evaluated by means of a performance measurement. In the context of training \acp{FFNN}, the underlying model is trained across a number of datasets. Each dataset is split into a training and test set. The training set is used to train the model, while the test set is used to evaluate the model's performance on unseen data. Training epochs are split into mini-batch iterations with a specified mini-batch size. Evaluation takes place at every mini-batch step. Due to the stochastic nature of the experiments, each experiment is repeated over a number of runs to provide sufficient sample for statistical certainty about the outcome of the results. Each run contains a different random seed, such that each run is unique. The evaluation data forms the results of the empirical test. These results are analysed for statistical significance, from which findings and conclusions are then made. The null hypothesis is then either rejected or verified based on these findings.

@@

This section provides the detail around the different datasets that are used throughout the empirical process. These datasets originate from the UCI Machine Learning Repository~\cite{ref:uci:2022}. Datasets are grouped by problem type and include 8 classification and 7 regression datasets. The classification datasets are given in Table \ref{tab:methodology:datasets:classification} and the regression datasets are given in Table \ref{tab:methodology:datasets:regression} below.

\begin{table}[htbp]
      \centering
      \caption{Classification datasets}
      \label{tab:methodology:datasets:classification}%
      \par\bigskip
      \resizebox{\textwidth}{!}{
            \begin{tabular}{ccccccccc}
                  \textbf{dataset} & \textbf{output} & \textbf{types}             & \textbf{attributes} & \textbf{classes} & \textbf{instances} & \textbf{batch} & \textbf{steps} & \textbf{citation}          \\
                  \midrule
                  iris             & multivariate    & real                       & 4                   & 3                & 150                & 16             & 10             & ~\cite{ref:fisher:1936}    \\
                  car              & multivariate    & categorical                & 6                   & 4                & 1728               & 128            & 14             & ~\cite{ref:bohanec:1988}   \\
                  abalone          & multivariate    & categorical, integer, real & 8                   & 28               & 4177               & 256            & 17             & ~\cite{ref:waugh:1995}     \\
                  wine quality     & multivariate    & real                       & 12                  & 11               & 4898               & 256            & 20             & ~\cite{ref:cortez:2009}    \\
                  mushroom         & multivariate    & categorical                & 22                  & 2                & 8214               & 512            & 17             & ~\cite{ref:schlimmer:1987} \\
                  bank             & multivariate    & real                       & 17                  & 2                & 45211              & 512            & 89             & ~\cite{ref:moro:2014}      \\
                  adult            & multivariate    & categorical, integer       & 14                  & 2                & 48842              & 256            & 191            & ~\cite{ref:kohavi:1996}    \\
                  diabetic         & multivariate    & integer                    & 55                  & 3                & 100000             & 1024           & 98             & ~\cite{ref:strack:2014}    \\
            \end{tabular}%
      }
\end{table}%

\begin{table}[htbp]
      \centering
      \caption{Regression datasets}
      \label{tab:methodology:datasets:regression}%
      \par\bigskip
      \resizebox{\textwidth}{!}{
            \begin{tabular}{cccccccc}
                  \textbf{dataset}    & \textbf{output}          & \textbf{types} & \textbf{attributes} & \textbf{instances} & \textbf{batch} & \textbf{steps} & \textbf{citation}         \\
                  \midrule
                  fish toxicity       & multivariate             & real           & 7                   & 908                & 64             & 15             & ~\cite{ref:cassotti:2015} \\
                  housing             & univariate               & real           & 13                  & 506                & 32             & 16             & ~\cite{ref:harrison:1978} \\
                  forest fires        & multivariate             & real           & 13                  & 517                & 32             & 17             & ~\cite{ref:cortez:2007}   \\
                  student performance & multivariate             & integer        & 33                  & 649                & 32             & 21             & ~\cite{ref:cortez:2008}   \\
                  parkinsons          & multivariate             & integer, real  & 26                  & 5875               & 256            & 23             & ~\cite{ref:tsanas:2009}   \\
                  air quality         & multivariate, timeseries & real           & 15                  & 9358               & 256            & 37             & ~\cite{ref:de:2008}       \\
                  bike                & univariate               & integer, real  & 16                  & 17389              & 256            & 68             & ~\cite{ref:fanaee:2014}   \\
            \end{tabular}%
      }
\end{table}%

Details on how the datasets where preprocessed and prepared is given in Appendix \ref{app:datasets}. The concept of class balancing is briefly discussed next.

@@

A number of classification datasets given in Table \ref{tab:methodology:datasets:classification} contain an unbalanced representation of classes. The empirical process defined in this dissertation does not apply mechanisms to cater for class balancing, in order to eliminate as many variables and factors in the empirical process as possible, as it is already a complicated process.  It is therefore suggested that the \acs{BHH} first be studied under the assumption of balanced classes, before applying class balancing techniques. In future research opportunities, class balancing should be utilised.

The next section presents the \acp{FFNN}/models and the configurations that are used in the empirical process.

@@

All models that are trained in this dissertation follow implementations of shallow \acp{FFNN}, meaning they only have one hidden layer. The architecture of a model is dependent on the dataset being trained on, the type of optimisation problem, the number of input dimensions and the number of output dimensions. Models' initial weights are initialised by means of \index{Glorot uniform sampling}Glorot uniform sampling. The models and their configuration, as it is used for each dataset, is given in Table \ref{tab:methodology:models:configurations} below.

% Table generated by Excel2LaTeX from sheet 'Model Configurations'
\begin{table}[htbp]
      \centering
      \caption{Model configurations}
      \label{tab:methodology:models:configurations}%
      \par\bigskip
      \resizebox{\textwidth}{!}{
            \begin{tabular}{rcccccccc}
                  \textbf{dataset}    & \textbf{inputs} & \textbf{hidden} & \textbf{output} & \textbf{biases} & \textbf{parameters} & \textbf{topology} & \textbf{l1 activation} & \textbf{l2 activation} \\
                  \midrule
                  fish toxicity       & 6               & 3               & 1               & yes             & 25                  & dense             & LReLU ($\alpha = 0.3$) & sigmoid                \\
                  iris                & 4               & 5               & 3               & yes             & 43                  & dense             & LReLU ($\alpha = 0.3$) & softmax                \\
                  air quality         & 12              & 8               & 1               & yes             & 113                 & dense             & LReLU ($\alpha = 0.3$) & sigmoid                \\
                  housing             & 13              & 8               & 1               & yes             & 121                 & dense             & LReLU ($\alpha = 0.3$) & sigmoid                \\
                  wine quality        & 13              & 10              & 7               & yes             & 217                 & dense             & LReLU ($\alpha = 0.3$) & softmax                \\
                  parkinsons          & 21              & 10              & 1               & yes             & 231                 & dense             & LReLU ($\alpha = 0.3$) & sigmoid                \\
                  car                 & 21              & 10              & 4               & yes             & 264                 & dense             & LReLU ($\alpha = 0.3$) & softmax                \\
                  forest fires        & 43              & 16              & 1               & yes             & 721                 & dense             & LReLU ($\alpha = 0.3$) & sigmoid                \\
                  abalone             & 10              & 36              & 28              & yes             & 1432                & dense             & LReLU ($\alpha = 0.3$) & softmax                \\
                  bank                & 51              & 32              & 1               & yes             & 1697                & dense             & LReLU ($\alpha = 0.3$) & softmax                \\
                  bike                & 61              & 32              & 1               & yes             & 2017                & dense             & LReLU ($\alpha = 0.3$) & sigmoid                \\
                  student performance & 99              & 32              & 1               & yes             & 3233                & dense             & LReLU ($\alpha = 0.3$) & sigmoid                \\
                  adult               & 108             & 64              & 1               & yes             & 7041                & dense             & LReLU ($\alpha = 0.3$) & softmax                \\
                  mushroom            & 117             & 64              & 1               & yes             & 7617                & dense             & LReLU ($\alpha = 0.3$) & softmax                \\
                  diabetic            & 2369            & 32              & 3               & yes             & 75939               & dense             & LReLU ($\alpha = 0.3$) & softmax                \\
            \end{tabular}%
      }
\end{table}%

@@

            \begin{tabular}{llll}
                  \textbf{heuristic} & \textbf{configuration}    & \textbf{value} & \textbf{citation}          \\
                  \midrule
                  sgd                & learning rate             & 0.1 (0.01)     & ~\cite{ref:sutskever:2013} \\
                  momentum           & learning rate             & 0.1 (0.01)     & ~\cite{ref:sutskever:2013} \\
                                     & momentum                  & 0.9            &                            \\
                  nag                & learning rate             & 0.1 (0.01)     & ~\cite{ref:sutskever:2013} \\
                                     & momentum                  & 0.9            &                            \\
                  adagrad            & learning rate             & 0.1 (0.01)     & ~\cite{ref:duchi:2011}     \\
                                     & epsilon                   & 1E-07          &                            \\
                  rmsprop            & learning rate             & 0.1 (0.01)     & ~\cite{ref:hinton:2012}    \\
                                     & rho                       & 0.95           &                            \\
                                     & epsilon                   & 1E-07          &                            \\
                  adadelta           & learning rate             & 1.0 (0.95)     & ~\cite{ref:zeiler:2012}    \\
                                     & rho                       & 0.95           &                            \\
                                     & epsilon                   & 0.0000001      &                            \\
                  adam               & learning rate             & 0.1 (0.01)     & ~\cite{ref:kingma:2014}    \\
                                     & beta1                     & 0.9            &                            \\
                                     & beta2                     & 0.95           &                            \\
                                     & epsilon                   & 1E-07          &                            \\
                  pso                & population size           & 10             & ~\cite{ref:van:2010}       \\
                                     & learning rate             & 0.1 (0.01)     &                            \\
                                     & inertia weight (w)        & 0.729844       &                            \\
                                     & cognitive control (c1)    & 1.49618        &                            \\
                                     & social control (c2)       & 1.49618        &                            \\
                                     & velocity clip min         & -1.0           &                            \\
                                     & velocity clip max         & 1.0            &                            \\
                  de                 & population size           & 10             & ~\cite{ref:mezura:2006}    \\
                                     & selection strategy        & best           &                            \\
                                     & xo strategy               & exp            &                            \\
                                     & recombination probability & 0.9 (0.1)      &                            \\
                                     & beta                      & 2.0 (0.1)      &                            \\
                  ga                 & population size           & 10             & ~\cite{ref:lambora:2019}   \\
                                     & selection strategy        & rand           &                            \\
                                     & xo strategy               & bin            &                            \\
                                     & mutation rate             & 0.2 (0.05)     &                            \\
            \end{tabular}%
      }
\end{table}%

Note that from Table \ref{tab:methodology:heuristics}, values that are configured to make use of a decay schedule are presented with the initial value first and the decay rate in brackets next to it. The number of steps is the total number of mini-batch training steps that are executed for that particular dataset.

Furthermore, it should be noted that a learning rate was added to \acs{PSO} as an attempt to avoid overshooting solutions later in the training process. This parameter does not traditionally form part of \acs{PSO}, but was found to be useful.

Section \ref{sec:bhh:heuristic:proxies} presented the concept of a mapping of proxied \index{heuristic}heuristic state update operations. The mapping of proxied \index{heuristic}heuristic state update operations, implemented by the \acs{BHH} in the empirical process, is given in Figure \ref{fig:methodology:heuristics:proxies}.

@@

In Figure \ref{fig:methodology:heuristics:proxies}, cells containing \textbf{x}, indicate that the associated \index{heuristic}heuristic implements that particular state parameter explicitly, and cells containing \textbf{o} indicate that the state parameter is implicitly implemented as part of the \acs{BHH} algorithm. The required mapping of proxied \index{heuristic}heuristic state operations is then implemented as a lookup of the table presented in Figure \ref{fig:methodology:heuristics:proxies}.

The following section provides the details around the \acs{BHH} baseline configuration.

@@

% Table generated by Excel2LaTeX from sheet 'BHH Baseline'
\begin{table}[htbp]
      \centering
      \caption{The \acs{BHH} baseline configuration as it is used in the empirical study.}
      \label{tab:methodology:bhh_baseline_configuration}%
      \par\bigskip
      \resizebox{0.65\textwidth}{!}{
            \begin{tabular}{rrcc}
                  \multicolumn{1}{c}{\textbf{hyper-heuristic}} & \multicolumn{1}{c}{\textbf{variant}} & \textbf{configuration} & \textbf{value} \\
                  \midrule
                  \multicolumn{1}{l}{bhh}                      & \multicolumn{1}{l}{baseline}         & heuristic pool         & all            \\
                                                               &                                      & population             & 5              \\
                                                               &                                      & credit                 & ibest          \\
                                                               &                                      & burn in                & 0              \\
                                                               &                                      & reselection            & 10             \\
                                                               &                                      & replay                 & 10             \\
                                                               &                                      & reanalysis             & 10             \\
                                                               &                                      & normalise              & false          \\
                                                               &                                      & discounted rewards     & false          \\
            \end{tabular}%
      }
\end{table}%

@@

The following section provides the performance measurements that are implemented.

@@

After training has been completed, the \textit{average rank}, based on test loss, for all configurations, is calculated at each mini-batch step across all runs. The test loss metric and average rank are statistically analysed and used to derive findings and conclusions.

The following section provides the stopping conditions that are used.

@@

The following sections provide the various experimental groups that are formulated and empirically tested.

@@

This section presents the experimental groups that are executed in the empirical process. Three main experimental groups are formulated. These include a behavioural case study of the \acs{BHH} as it relates to an example problem, a critical evaluation of the \acs{BHH}'s baseline performance in comparison to other standalone low-level \index{heuristic}heuristics, and finally an analysis of different \acs{BHH} variants is done by analysing the effects of different hyper-parameter values on the outcomes of the \acs{BHH}. Each of these are discussed in more detail in the following sections.

@@

The behavioural case study is concerned with studying the behaviour of the \acs{BHH} baseline configuration as it relates to an example problem dataset, across a number of runs. The example problem dataset is the iris dataset, presented in Table \ref{tab:methodology:datasets:classification}. The behavioural case study is meant to provide an introductory analysis of the inner workings of the \acs{BHH} and includes analysis of the selection mechanism, as well as the perturbative nature of the \acs{BHH}. The behavioural case study is not meant to be statistically analysed, but to rather provide a detailed visual analysis of a number of hand-picked example runs, to determine if the \acs{BHH} is behaving as expected. From these observations, it is determined if the \acs{BHH} is learning and that selection is indeed biasing towards better performance. These observations also provide an opportunity to observe the outcome of the perturbative nature of the \acs{BHH}, which includes proxied \index{heuristic}heuristic update step operations.

@@

Additional to the \acs{BHH} baseline configuration, two more \acs{BHH} configurations are included. These include \acs{BHH} configurations where the \index{heuristic pool}heuristic pool only makes use of gradient-based \index{heuristic}heuristics, and a configuration where the \index{heuristic pool}heuristic pool only makes use of meta-heuristics. The intent behind the inclusion of these configurations is to determine the effectiveness of multi-method approaches in the \index{heuristic pool}heuristic pool as it applies to training \acp{FFNN}.

@@

\begin{table}[htbp]
      \centering
      \caption{BHH variants and their configuration}
      \label{tab:methodology:experiments:bhh_variants}%
      \par\bigskip
      \resizebox{0.9\textwidth}{!}{
            \begin{tabular}{rcc}
                  \multicolumn{1}{c}{\textbf{hyper-heuristic}} & \textbf{variant}   & \textbf{values}                       \\
                  \midrule
                  \multicolumn{1}{l}{bhh}                      & heuristic pool     & all, gd, mh                           \\
                                                               & population         & 5, 10, 15, 20, 25                     \\
                                                               & credit             & ibest, pbest, rbest, gbest, symmetric \\
                                                               & reselection        & 1, 5, 10, 15, 20                      \\
                                                               & replay             & 1, 5, 10, 15, 20                      \\
                                                               & reanalysis         & 1, 5, 10, 15, 20                      \\
                                                               & burn in            & 0, 5, 10, 15, 20                      \\
                                                               & normalise          & false, true                           \\
                                                               & discounted rewards & false, true                           \\
            \end{tabular}%
      }
\end{table}%

In Table \ref{tab:methodology:experiments:bhh_variants}, the \index{heuristic pool}heuristic pool configuration \textit{gd} refers to the \index{heuristic pool}heuristic pool configuration where only gradient-based \index{heuristic}heuristics are included, and \textit{mh} refers to the \index{heuristic pool}heuristic pool configuration where only meta-heuristics are included.

The next section presents the method that is followed for statistical analysis.

@@

The evaluation of outcome is based on the aggregated statistical results as mentioned above. A descriptive analysis is executed on the spread of the data. The results are analysed and checked for overfitting and outliers are identified. The skewness of the results is evaluated per dataset and the \index{Shapiro-Wilk}Shapiro-Wilk test for normality ($\alpha$ = 0.001) is used to determine if the results are normally distributed. Furthermore, the \index{Levene}Levene's test for equality of variance ($\alpha$ = 0.001) is used. The output of the full statistical analysis is presented in Appendix \ref{app:statistical_analysis}.

Dependent on the outcomes of the above statistical tests, the appropriate statistical significance test is then executed. For experiments where there are only two configurations, the Mann-Whitney U\index{Mann-Whitney U} independent samples t-Test ($\alpha$ = 0.001) is executed. For experiments with three or more configurations, the \acf{ANOVA} statistical test ($\alpha$ = 0.001) is used. The\index{Kruskal-Wallis} Kruskal-Wallis ranked non-parametric test for statistical significance ($\alpha$ = 0.001) is used for cases where data is not normally distributed.

Regardless of the statistical test that is used, a post-hoc Tukey honest significant difference test ($\alpha$ = 0.001) is used from which significant ranking is retrieved. Descriptive and critical difference plots are then retrieved from these results to provide visual aid.

@@

All experiments were run on the Centre for High Performance Computing (CHPC) cluster. A total of 14 different server were used, each running 24-56 cores and 256GB of memory. The entire empirical process took 6 days.

>>>>

Each empirical test starts with a question to be answered. A hypothesis is formulated for the outcome of the empirical test. The empirical tests are then designed around the implementation of the components under evaluation. Each empirical test defines the configuration of elements and generally include a set of parameters that are altered between experiments. Each experiment is evaluated by means of a performance measurement. In the context of training \acp{FFNN}, the underlying model is trained across a number of datasets. Each dataset is split into a training set comprising of 80\% of the data, and a test set comprising of 20\% of the data. The training set is used to train the model, while the test set is used to evaluate the model's performance on unseen data. Training epochs are split into mini-batch iterations with a specified mini-batch size per dataset. Evaluation takes place at every mini-batch step. Due to the stochastic nature of the experiments, each experiment is repeated over a number of runs to provide sufficient sample for statistical certainty about the outcome of the results. Each run contains a different random seed, such that each run is unique. The evaluation data forms the results of the empirical test. These results are analysed for statistical significance, from which findings and conclusions are then made. The null hypothesis is then either rejected or verified based on these findings.

@@

This section provides the detail around the different datasets that are used throughout the empirical process. These datasets originate from the UCI Machine Learning Repository~\cite{ref:uci:2022}. Datasets are grouped by problem type and include seven classification and seven regression datasets. The classification datasets are given in Table \ref{tab:methodology:datasets:classification} and the regression datasets are given in Table \ref{tab:methodology:datasets:regression} below.

\begin{table}[htb]
      \centering
      \caption{Classification datasets}
      \label{tab:methodology:datasets:classification}%
      \par\bigskip
      \resizebox{\textwidth}{!}{
            \begin{tabular}{ccccccccc}
                  \textbf{dataset} & \textbf{output} & \textbf{types}             & \textbf{attributes} & \textbf{classes} & \textbf{instances} & \textbf{batch} & \textbf{steps} & \textbf{citation}          \\
                  \midrule
                  iris             & multivariate    & real                       & 4                   & 3                & 150                & 16             & 10             & ~\cite{ref:fisher:1936}    \\
                  car              & multivariate    & categorical                & 6                   & 4                & 1728               & 128            & 14             & ~\cite{ref:bohanec:1988}   \\
                  abalone          & multivariate    & categorical, integer, real & 8                   & 28               & 4177               & 256            & 17             & ~\cite{ref:waugh:1995}     \\
                  wine quality     & multivariate    & real                       & 12                  & 11               & 4898               & 256            & 20             & ~\cite{ref:cortez:2009}    \\
                  mushroom         & multivariate    & categorical                & 22                  & 2                & 8214               & 512            & 17             & ~\cite{ref:schlimmer:1987} \\
                  bank             & multivariate    & real                       & 17                  & 2                & 45211              & 512            & 89             & ~\cite{ref:moro:2014}      \\
                  diabetic         & multivariate    & integer                    & 55                  & 3                & 100000             & 1024           & 98             & ~\cite{ref:strack:2014}    \\
            \end{tabular}%
      }
\end{table}%

\begin{table}[htb]
      \centering
      \caption{Regression datasets}
      \label{tab:methodology:datasets:regression}%
      \par\bigskip
      \resizebox{\textwidth}{!}{
            \begin{tabular}{cccccccc}
                  \textbf{dataset}    & \textbf{output}          & \textbf{types} & \textbf{attributes} & \textbf{instances} & \textbf{batch} & \textbf{steps} & \textbf{citation}         \\
                  \midrule
                  fish toxicity       & multivariate             & real           & 7                   & 908                & 64             & 15             & ~\cite{ref:cassotti:2015} \\
                  housing             & univariate               & real           & 13                  & 506                & 32             & 16             & ~\cite{ref:harrison:1978} \\
                  forest fires        & multivariate             & real           & 13                  & 517                & 32             & 17             & ~\cite{ref:cortez:2007}   \\
                  student performance & multivariate             & integer        & 33                  & 649                & 32             & 21             & ~\cite{ref:cortez:2008}   \\
                  parkinsons          & multivariate             & integer, real  & 26                  & 5875               & 256            & 23             & ~\cite{ref:tsanas:2009}   \\
                  air quality         & multivariate, timeseries & real           & 15                  & 9358               & 256            & 37             & ~\cite{ref:de:2008}       \\
                  bike                & univariate               & integer, real  & 16                  & 17389              & 256            & 68             & ~\cite{ref:fanaee:2014}   \\
            \end{tabular}%
      }
\end{table}%

Details on how the datasets where preprocessed and prepared is given in Appendix \ref{app:datasets}.

@@

A number of classification datasets given in Table \ref{tab:methodology:datasets:classification} contain an unbalanced representation of classes. The empirical process defined in this dissertation does not apply mechanisms to cater for class balancing, in order to eliminate as many variables and factors in the empirical process as possible. It is therefore suggested that the \acs{BHH} first be studied under the assumption of balanced classes, before applying class balancing techniques. In future research opportunities, class balancing should be utilised.

@@

All models that are trained in this dissertation follow implementations of shallow \acp{FFNN}, meaning that they only have one hidden layer. The architecture of a model is dependent on the dataset being trained on, the type of optimisation problem, the number of input dimensions and the number of output dimensions. The number of hidden layers were determined empirically. Weights are initialised by means of \index{Glorot uniform sampling}Glorot uniform sampling~\cite{ref:glorot:2010}. The models and their configuration, as it is used for each dataset, are given in Table \ref{tab:methodology:models:configurations}.

\begin{table}[htb]
      \centering
      \caption{Model configurations}
      \label{tab:methodology:models:configurations}%
      \par\bigskip
      \resizebox{\textwidth}{!}{
            \begin{tabular}{rcccccccc}
                  \textbf{dataset}    & \textbf{inputs} & \textbf{hidden} & \textbf{output} & \textbf{biases} & \textbf{parameters} & \textbf{topology} & \textbf{l1 activation} & \textbf{l2 activation} \\
                  \midrule
                  fish toxicity       & 6               & 3               & 1               & yes             & 25                  & dense             & LReLU ($\alpha = 0.3$) & sigmoid                \\
                  iris                & 4               & 5               & 3               & yes             & 43                  & dense             & LReLU ($\alpha = 0.3$) & softmax                \\
                  air quality         & 12              & 8               & 1               & yes             & 113                 & dense             & LReLU ($\alpha = 0.3$) & sigmoid                \\
                  housing             & 13              & 8               & 1               & yes             & 121                 & dense             & LReLU ($\alpha = 0.3$) & sigmoid                \\
                  wine quality        & 13              & 10              & 7               & yes             & 217                 & dense             & LReLU ($\alpha = 0.3$) & softmax                \\
                  parkinsons          & 21              & 10              & 1               & yes             & 231                 & dense             & LReLU ($\alpha = 0.3$) & sigmoid                \\
                  car                 & 21              & 10              & 4               & yes             & 264                 & dense             & LReLU ($\alpha = 0.3$) & softmax                \\
                  forest fires        & 43              & 16              & 1               & yes             & 721                 & dense             & LReLU ($\alpha = 0.3$) & sigmoid                \\
                  abalone             & 10              & 36              & 28              & yes             & 1432                & dense             & LReLU ($\alpha = 0.3$) & softmax                \\
                  bank                & 51              & 32              & 1               & yes             & 1697                & dense             & LReLU ($\alpha = 0.3$) & softmax                \\
                  bike                & 61              & 32              & 1               & yes             & 2017                & dense             & LReLU ($\alpha = 0.3$) & sigmoid                \\
                  student performance & 99              & 32              & 1               & yes             & 3233                & dense             & LReLU ($\alpha = 0.3$) & sigmoid                \\
                  adult               & 108             & 64              & 1               & yes             & 7041                & dense             & LReLU ($\alpha = 0.3$) & softmax                \\
                  mushroom            & 117             & 64              & 1               & yes             & 7617                & dense             & LReLU ($\alpha = 0.3$) & softmax                \\
                  diabetic            & 2369            & 32              & 3               & yes             & 75939               & dense             & LReLU ($\alpha = 0.3$) & softmax                \\
            \end{tabular}%
      }
\end{table}%

@@

            \begin{tabular}{llll}
                  \textbf{heuristic} & \textbf{configuration}    & \textbf{value} & \textbf{citation}          \\
                  \midrule
                  sgd                & learning rate             & 0.1 (0.01)     & ~\cite{ref:sutskever:2013} \\
                  momentum           & learning rate             & 0.1 (0.01)     & ~\cite{ref:sutskever:2013} \\
                                     & momentum                  & 0.9            &                            \\
                  nag                & learning rate             & 0.1 (0.01)     & ~\cite{ref:sutskever:2013} \\
                                     & momentum                  & 0.9            &                            \\
                  adagrad            & learning rate             & 0.1 (0.01)     & ~\cite{ref:duchi:2011}     \\
                                     & epsilon                   & 1E-07          &                            \\
                  rmsprop            & learning rate             & 0.1 (0.01)     & ~\cite{ref:hinton:2012}    \\
                                     & rho                       & 0.95           &                            \\
                                     & epsilon                   & 1E-07          &                            \\
                  adadelta           & learning rate             & 1.0 (0.95)     & ~\cite{ref:zeiler:2012}    \\
                                     & rho                       & 0.95           &                            \\
                                     & epsilon                   & 1E-07          &                            \\
                  adam               & learning rate             & 0.1 (0.01)     & ~\cite{ref:kingma:2014}    \\
                                     & beta1                     & 0.9            &                            \\
                                     & beta2                     & 0.999          &                            \\
                                     & epsilon                   & 1E-07          &                            \\
                  pso                & population size           & 10             & ~\cite{ref:van:2010}       \\
                                     & learning rate             & 1.0 (0.9)      &                            \\
                                     & inertia weight (w)        & 0.729844       &                            \\
                                     & cognitive control (c1)    & 1.49618        &                            \\
                                     & social control (c2)       & 1.49618        &                            \\
                                     & velocity clip min         & -1.0           &                            \\
                                     & velocity clip max         & 1.0            &                            \\
                  de                 & population size           & 10             & ~\cite{ref:mezura:2006}    \\
                                     & selection strategy        & best           &                            \\
                                     & xo strategy               & exp            &                            \\
                                     & recombination probability & 0.9 (0.1)      &                            \\
                                     & beta                      & 2.0 (0.1)      &                            \\
                  ga                 & population size           & 10             & ~\cite{ref:lambora:2019}   \\
                                     & selection strategy        & rand           &                            \\
                                     & xo strategy               & bin            &                            \\
                                     & mutation rate             & 0.2 (0.05)     &                            \\
            \end{tabular}%
      }
\end{table}%

Note from Table \ref{tab:methodology:heuristics} that values that are configured to make use of a decay schedule are presented with the initial value first and the decay rate in brackets next to it. The number of steps is the total number of mini-batch training steps that are executed for that particular dataset.

Furthermore, it should be noted that a learning rate was added to \acs{PSO} as an attempt to avoid overshooting solutions later in the training process. A learning rate parameter does not traditionally form part of \acs{PSO}, but was found to be useful. The position update step for \acs{PSO} using a learning rate is then defined as

\begin{equation}
      \label{eq:heuristics:pso:position_learning_rate}
      \begin{split}
            x_{ij}(t+1) = x_{ij}(t) + \eta v_{ij}(t+1)
      \end{split}
\end{equation}

where $\eta$ is the added learning rate with $0 \leq \eta \leq 1, \eta \in \mathbb{R}$.

Section \ref{sec:bhh:heuristic:proxies} presented the concept of a mapping of proxied \index{heuristic}heuristic state update operations. The mapping of proxied \index{heuristic}heuristic state update operations implemented by the \acs{BHH} in the empirical process is given in Figure \ref{fig:methodology:heuristics:proxies}.

@@

In Figure \ref{fig:methodology:heuristics:proxies}, cells containing \textbf{x} indicate that the associated \index{heuristic}heuristic implements that particular state parameter explicitly, and cells containing \textbf{o} indicate that the state parameter is implicitly implemented as part of the \acs{BHH} algorithm. The required mapping of proxied \index{heuristic}heuristic state operations is then implemented as a lookup of the table presented in Figure \ref{fig:methodology:heuristics:proxies}.

@@

% Table generated by Excel2LaTeX from sheet 'BHH Baseline'
\begin{table}[htb]
      \centering
      \caption{The \acs{BHH} baseline configuration as it is used in the empirical study.}
      \label{tab:methodology:bhh_baseline_configuration}%
      \par\bigskip
      \resizebox{0.65\textwidth}{!}{
            \begin{tabular}{rrcc}
                  \multicolumn{1}{c}{\textbf{hyper-heuristic}} & \multicolumn{1}{c}{\textbf{variant}} & \textbf{configuration} & \textbf{value} \\
                  \midrule
                  \multicolumn{1}{l}{bhh}                      & \multicolumn{1}{l}{baseline}         & heuristic pool         & all            \\
                                                               &                                      & population             & 5              \\
                                                               &                                      & credit                 & ibest          \\
                                                               &                                      & burn in                & 0              \\
                                                               &                                      & reselection            & 10             \\
                                                               &                                      & replay                 & 10             \\
                                                               &                                      & reanalysis             & 10             \\
                                                               &                                      & normalise              & false          \\
                                                               &                                      & discounted rewards     & false          \\
            \end{tabular}%
      }
\end{table}%

@@

@@

The test set is used as a validation set during training. All performance metrics are measured for the training set as well as the test/validation set during training. As such, a divergence in performance metrics between the training and test/validation set is used to evaluate for overfitting during the training process.

After training has been completed, the \textit{average rank}, based on test loss, for all configurations, is calculated at each mini-batch step. A total of 30 independent runs with different random seed configurations are executed for each empirical test. The test loss metric and average rank are then statistically analysed across all runs, and are used to derive findings and conclusions.

@@

@@

This section presents the experimental groups that are executed in the empirical process. Three main experimental groups are formulated. Each of these are discussed in more detail in the following sections.

@@

The behavioural case study analyses the behaviour of the \acs{BHH} baseline configuration as it relates to an example problem dataset, across 30 independent runs. The example problem dataset is the iris dataset, presented in Table \ref{tab:methodology:datasets:classification}. The behavioural case study is meant to provide an introductory analysis of the inner workings of the \acs{BHH} and includes analysis of the selection mechanism, as well as the perturbative nature of the \acs{BHH}. The behavioural case study is not meant to be statistically analysed, but to rather provide a detailed visual analysis of a number of hand-picked example runs, to determine if the \acs{BHH} is behaving as expected. From these observations, it is determined if the \acs{BHH} is learning and that selection is indeed biasing towards better performance. These observations also provide an opportunity to observe the outcome of the perturbative nature of the \acs{BHH}, which includes proxied \index{heuristic}heuristic update step operations.

@@

Additional to the \acs{BHH} baseline configuration, two more \acs{BHH} configurations are included. These include \acs{BHH} configurations where the \index{heuristic pool}heuristic pool only makes use of gradient-based \index{heuristic}heuristics, and a configuration where the \index{heuristic pool}heuristic pool only makes use of \acp{MH}. The intent behind the inclusion of these configurations is to determine the effectiveness of multi-method approaches in the \index{heuristic pool}heuristic pool as it applies to training \acp{FFNN}.

@@

\begin{table}[htb]
      \centering
      \caption{BHH variants and their configuration}
      \label{tab:methodology:experiments:bhh_variants}%
      \par\bigskip
      \resizebox{0.9\textwidth}{!}{
            \begin{tabular}{rcc}
                  \multicolumn{1}{c}{\textbf{hyper-heuristic}} & \textbf{variant}   & \textbf{values}                       \\
                  \midrule
                  \multicolumn{1}{l}{bhh}                      & heuristic pool     & all, gd, mh                           \\
                                                               & population         & 5, 10, 15, 20, 25                     \\
                                                               & credit             & ibest, pbest, rbest, gbest, symmetric \\
                                                               & reselection        & 1, 5, 10, 15, 20                      \\
                                                               & replay             & 1, 5, 10, 15, 20                      \\
                                                               & reanalysis         & 1, 5, 10, 15, 20                      \\
                                                               & burn in            & 0, 5, 10, 15, 20                      \\
                                                               & normalise          & false, true                           \\
                                                               & discounted rewards & false, true                           \\
            \end{tabular}%
      }
\end{table}%

In Table \ref{tab:methodology:experiments:bhh_variants}, the \index{heuristic pool}heuristic pool configuration \textit{gd} refers to the \index{heuristic pool}heuristic pool configuration where only gradient-based \index{heuristic}heuristics are included, and \textit{mh} refers to the \index{heuristic pool}heuristic pool configuration where only \acp{MH} are included.

@@

The evaluation of outcome is based on the aggregated statistical results as mentioned above. A descriptive analysis is executed on the spread of the data. The results are analysed and checked for overfitting and outliers are identified. The skewness of the results is evaluated per dataset and the \index{Shapiro-Wilk}Shapiro-Wilk test for normality~\cite{ref:shapiro:1965} ($\alpha$ = 0.001) is used to determine if the results are normally distributed. Furthermore, the \index{Levene}Levene's test for equality of variance~\cite{ref:levene:1961} ($\alpha$ = 0.001) is used. The output of the full statistical analysis is presented in Appendix \ref{app:statistical_analysis}.

Dependent on the outcomes of the above statistical tests, the appropriate statistical significance test is then executed. For experiments where there are only two configurations, the Mann-Whitney U\index{Mann-Whitney U} independent samples t-Test~\cite{ref:mann:1947} ($\alpha$ = 0.001) is executed. For experiments with three or more configurations, the \acs{ANOVA} statistical test~\cite{ref:fisher:1921} ($\alpha$ = 0.001) is used. The\index{Kruskal-Wallis} Kruskal-Wallis ranked non-parametric test~\cite{ref:kruskal:1952} for statistical significance ($\alpha$ = 0.001) is used for cases where data is not normally distributed.

Regardless of the statistical test that is used, a post-hoc Tukey honest significant difference test~\cite{ref:tukey:1949} ($\alpha$ = 0.001) is used from which significant ranking is retrieved. Descriptive and critical difference plots are then retrieved from these results to provide visual aid.

@@

All experiments were run on the Centre for High Performance Computing (CHPC) cluster. A total of 14 different server were used, each running 24 to 56 cores and 256GB of memory. The entire empirical process took six days.
